Tree biodiversity and ecosystem function relationship across southern African woodlands

### 2020-08-18 manuscript/regional_befr_manuscript.tex
Removed cache

\title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
\author{}
\date{}

\input{include/n_plots}
\input{include/hull_cover}
\input{include/path_coef_struc}
\input{include/path_coef_full}
\input{include/corr_coef}
\input{include/dens_stats}
\input{include/perc_small_agb.tex}

@@

{\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}}

@@

Authors: Godlee, J. L.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}

\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom \\
\textsuperscript{2}: Some other address

@@

School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom

@@

This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (J. L. Godlee, Grant No. NE/L002558/1). The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No. NE/P008755/1). We thank all data providers and the field assistance they received when collecting plot data.

@@

\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear. In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function. Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for gradients in stem density, resource availability, disturbance through fire, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

\textbf{Location:} Southern African savannas and woodlands

\textbf{Time period:} 2010-2019

@@

\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent . We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, and stem density.

\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, observed mainly via the increasing effect of woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity.

\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and ecosystem function of highly disturbed southern African savannas and woodlands. Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level. Biodiversity loss, predominantly through human actions in southern Africa, may have detrimental outcomes for ecosystem function, specifically woody biomass provision, which is critical in a region that relies on wood fuel as its principal domestic energy source.

@@

In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small-scale experimental studies conducted predominantly in grassland ecosystems, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}. Despite these concerted efforts, we continue to lack a nuanced, ecosystem agnostic understanding of the complex interactions between biodiversity, environment, and ecosystem function.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the nature of biotic activity within an ecosystem \citep{Jax2005}. This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported  relationship between biodiversity and ecosystem function invokes three main mechanisms to explain the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

Compared to forest ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire and herbivory, with African savannas possessing large herbivores absent from other savannas \citep{Sankaran2008, Levick2009}. Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{Dewees2010}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance may reduce stem density, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. Reduced stem density reduces competition between individuals, possibly weakening the effect of diversity on ecosystem function via niche complementarity. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects. This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.

The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEF relationship and its environmental determinants.

Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages.These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

A small number of studies in southern African woodlands, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and tree species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass. \citet{Sankaran2005} used data from 854 African woodland field sites to show that below a threshold of \textapprox{650 mm MAP}, precipitation sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.

Southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy, depending on their growth stages and species identity \citet{Solbrig1996}. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production. Occupancy of multiple canopy layers allows a fuller canopy with  greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas. Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.

High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results. We used Structural Equation Modelling (SEM) with path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass. Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the strength of the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), as affected by resource availability and disturbance, due to an increased importance of niche complementarity as stem density and therefore competition increases, and (3) tree species diversity will increase tree structural diversity (i.e. physiognomic diversity), which will provide an indirect path by which tree diversity increases woody biomass.

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealt{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion  \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.

Many plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009} were arranged in clusters of up to four 20x50 m plots, 20 metres apart. Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.

@@

We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads which are functionally tree-like. Woody lianas are scarce in our study plots and were not measured. Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Many plots in the dataset did not include data on stems <10 cm DBH. For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot level AGB.

All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level diversity estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice. For each tree, we recorded species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:

@@

where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective genera. For stems where tree height was unknown, the plots' climatic parameters, estimated from plot location, were used to estimate tree height, according to \citet{Chave2014}.

Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}. In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15). Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), soil organic carbon stocks (kg m\textsuperscript{-2}) soil sand content by volume (Sand g kg\textsuperscript{-1}) and soil nitrogen content (g kg\textsuperscript{-1}). These data are a modelled product derived from various remotely sensed and directly measured data sources. The degree of fire disturbance was calculated using the MODIS monthly burned area product at 0.01 degree resolution, 1.11 km at the equator (MCD64A1, \citealt{MODIS_burn}), as the number of observed fires between 2001 and 2018.

>>>>

\title{Tree biodiversity and ecosystem function relationship across southern African woodlands}
\author{}
\date{}

\input{include/n_plots}
\input{include/hull_cover}
\input{include/path_coef_struc}
\input{include/path_coef_full}
\input{include/corr_coef}
\input{include/dens_stats}
\input{include/perc_small_agb.tex}
\input{include/zam.tex}

@@

{\LARGE{Title: Tree biodiversity and ecosystem function relationship across southern African woodlands}}

@@

Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1,2}, Ryan, C. M.\textsuperscript{1}

\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
\textsuperscript{2}: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom

@@

School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom

@@

This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (J. L. Godlee, Grant No. NE/L002558/1). The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No. NE/P008755/1). We thank all data providers and the field assistance they received when collecting plot data. JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).

@@

\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear. It is posited that high levels of disturbance moderate the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function. Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for variation in stem density, resource availability, disturbance through fire, and vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

\textbf{Location:} Southern African savannas and woodlands
\textbf{Time period:} 2010-2019

@@

\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent. We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.

\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, across vegetation types, observed mainly via the increasing effect of woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}. Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.

\textbf{Main conclusions:} The study underlines the close association between tree diversity and ecosystem structure and function of highly disturbed southern African savannas and woodlands. Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.

@@

% BEFR has had previous global research
In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}. The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}. Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in artificial grassland mesocosms \citep{Tilman1994, Tilman2014}. Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, environment, and ecosystem function.

% Ecosystem function definition and mechanisms
Ecosystem functions can be defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}. This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resilience of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported BEF relationship invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

% Savannas are big and ecologically important
Savannas and woodlands are the dominant vegetation type across southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages. These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under international efforts to reduced deforestation and degradation (REDD+) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

% Savannas might not be the same as forests
Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. A threshold stem density may exist below which the effects of biodiversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.

% Savannas might no be the same as forests continued: facilitation effects
More diverse species assemblages may lead to facilitation effects between certain species combinations under the limiting environmental conditions prevalent across African savannas, such as low water availability. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects. This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.

% African savannas are understudied in this regard
The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on the BEF in southern African woodlands could greatly strengthen our understanding of a global BEF relationship and its environmental determinants. A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.

% \citep{Michaletz2014, Michaletz2018} \citep{Spasojevic2014}
%In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
% Due the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass. 
% Environment affects the BEF relationship, so must be accounted for.
Environmental variation is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}. Southern African woodlands particularly, occur over a wide range of precipitation, diurnal and annual temperature, and disturbance regimes \citep{Frost1996}. It is important therefore to account for environment and understand how it interacts with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass. \citet{Sankaran2005} and \citet{Lehmann2014} both report independently that total precipitation sets the upper limit for woody biomass in African savannas. \citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.

% Disturbance deserves another look
High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance-prone woodlands. If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

% Structural diversity is important too
Southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy, depending on their growth stages and species identity \citep{Solbrig1996}. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production. Occupancy of multiple canopy layers allows a fuller canopy with  greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas. Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.

% In this study :::
In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots spanning broad environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results. We used Structural Equation Modelling (SEM) with path analysis to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass. Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases. In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Situation
The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealt{White1987}). The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.

% Plots chosen
Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.

% Zambia disclaimer
\nzam{} plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009, Pelletier2018} were arranged in clusters of four 20x50 m plots, 20 metres apart. Data from each plot within a cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.

@@

% AGB
We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads, which are functionally tree-like. Woody lianas are scarce in our study plots and were not measured. Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Many plots in the dataset did not include data on stems <10 cm DBH. For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot level AGB.

% Number of stems
All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements. A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice. For each tree, we recorded species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.

% AGB Calculation
AGB for each plot (t ha\textsuperscript{-1}) was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:

@@

where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D$ is the DBH\textsubscript{e} (cm) at 1.3 m, and $H$ is the tree height (m). Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective genera. For stems where tree height was unknown, the plots' climatic parameters, estimated from plot location, were used to estimate tree height, according to \citet{Chave2014}.
